Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-11-15 manuscript/manuscript.tex

\author[1\authfn{1}*]{Christoph Wehmeyer}
\author[1\authfn{1}]{Martin K. Scherer}
\author[1\authfn{1}]{Tim Hempel}
\author[2]{Brooke E. Husic}
\author[1]{Simon Olsson}
\author[1*]{Frank Noé}
\affil[1]{Department of Mathematics and Computer Science, Freie Universität Berlin, Arnimallee 6, 14195 Berlin, Germany}
\affil[2]{Department of Chemistry, Stanford University, 333 Campus Drive, Stanford, California 94305, USA}

@@

In addition to publications on the theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
we also recommend the literature on TICA~\cite{tica,tica3,tica2,kinetic-maps},
transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,jhp-spectral-rate-theory,bhmm-preprint},
and variational techniques~\cite{noe-vac,vamp-preprint,gmrq},
as these topics play important roles within the standard MSM workflow.

@@

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,jhp-spectral-rate-theory,noe-proj-hid-msm,bhmm-preprint}.
HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space (illustrated in Fig.~\ref{fig:hmm-scheme}).
Instead, HMMs assume that there is an underlying (hidden) dynamic process which is Markovian
and gives rise to our observed data, e.g., the ($n$~states) discretized trajectories $s(t)$.
This is a powerful principle as we know that there is indeed an underlying process which is Markovian:
our molecular dynamics trajectories.

@@

@@

For an extensive discussion of details about HMM properties and the estimation algorithm in general, we suggest Ref.~\cite{hmm-tutorial}.
For its specific application to the discretization of MSMs using HMMs, we suggest Ref.~\cite{noe-proj-hid-msm}. A generalized extension for estimating this type of low-dimensional projection from the data is given in Ref.~\cite{wu2015projected}.

>>>>

\author[1\authfn{1}*]{Christoph Wehmeyer}
\author[1\authfn{1}]{Martin K. Scherer}
\author[1\authfn{1}]{Tim Hempel}
\author[1,2]{Brooke E. Husic}
\author[1]{Simon Olsson}
\author[1*]{Frank Noé}
\affil[1]{Department of Mathematics and Computer Science, Freie Universität Berlin, Arnimallee 6, 14195 Berlin, Germany}
\affil[2]{Department of Chemistry, Stanford University, 333 Campus Drive, Stanford, California 94305, USA}

@@

In addition to publications on the theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks,simon-amm},
we also recommend the literature on TICA~\cite{tica,tica3,tica2,kinetic-maps},
transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint},
and variational techniques~\cite{noe-vac,vamp-preprint,gmrq},
as these topics play important roles within the standard MSM workflow.

@@

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,noe-proj-hid-msm,bhmm-preprint}.
HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space (illustrated in Fig.~\ref{fig:hmm-scheme}).
Instead, HMMs assume that there is an underlying (hidden) dynamic process which is Markovian
and gives rise to our observed data, e.g., the ($n$~states) discretized trajectories $s(t)$.
This is a powerful principle as we know that there is indeed an underlying process which is Markovian:
our molecular dynamics trajectories.

@@

We can further compute a reversal of the emission matrix $\bm{\chi}\in\mathbb{R}^{m \times n}$:
the membership matrix $\mathbf{M}\in\mathbb{R}^{n \times m}$ which encodes
a fuzzy assignment of each of the $n$ observable microstates $s$ to the $m$ hidden states $\tilde{s}$ and,
thus, defines the \emph{coarse graining} of microstate.

@@
